Research Experience

### 2010-03-01 doc/thesis/thesis.tex
Merge ssh://oldcub.ath.cx/home/micahc/brainid

\begin{document}
\section*{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
neural activity. At the cost of rather limited temporal resolution, FMRI gives
high spatial for relatively cheap. Whereas other methods
of analyzing neural signals can be invasive or ill-posed, 
FMRI is relatively quick and easy, and its analysis straight forward. Because of
these benefits, FMRI has great potential for studying human cognition.
Unfortunately, FMRI does suffer from severe noise and, as previously
mentioned, low temporal resolution. Because of the extensive noise, 
and relative unpredictability of the absolute signal level, a large amount 
of conditioning is necessary to obtain "valid" activation. In
fact, typical activation detection does little to account for variations
in the shape of the signal, but instead is a direct T-test between the measured signal
and the expected shape of the signal. Because of these limitations,
multiple comparison tests are imperative (ref fish), and active regions
must be very large to be detected with any confidence. This paper
will introduce a more robust method of analyzing FMRI signals for 
activation, by means of estimating the BOLD parameters.

\section*{General Linear Model}

\section*{BOLD Model}

\subsection*{Preprocessing}
After FMRI data has been acquired it is always necessary to modify the
data in some way to make different runs comparable. Because FMRI signal
levels are essentially unit-less, at the very least it is necessary to convert
the data into \% difference from the baseline. Finding the baseline is the
first hurdle in dealing with FMRI. Various methods exist for doing this,
but they are all ad-hoc. It is common to treat each stimulus pulse independently,
ignoring the long decay of the BOLD signal. By treating each input separately,
the "base" for calculating the percent difference is merely the signal level 
before the stimulus is applied. The problem with this is that the BOLD signal
can be heavily delayed, and fall times can be as long as 10 or 15 seconds,
meaning that the assumption is clearly a false one. Another popular method 
of dealing with this problem, is to put all the time series through a high
pass filter. This will of course remove the DC component of the signal, and 
some amount of so called "drift". The problem with this method is that it is
not adaptive to the input. Huge variations in drift frequencies can exist 
in a single time-series. Thus, a single cutoff frequency could miss a significant
drift component, or it could remove \emph{actual} signal, if the cutoff frequency is
set too high.

\subsection*{

\section*{Introduction to Modeling}
The short term goal of current neurological models is to
learn something about the system being modeled.
While control (for instance in prosthetics), or predication
(such as BlueBrain) may some day be feasible, in the
short run we still have a great deal to learn about how our
brain works. By learning about specific systems (the dynamics
of a single person's brain), we may be able to diagnose diseases,
find their causes and even someday rectify the cause. Of course
by learning about various specific systems we can also learn 
about the general way our brains work. The reason then, for 
studying the BOLD model then is
to be able to derive the various system parameters and understand
why one brain region (say a voxel) reacts in one way, when another
acts completely differently.

While the General Linear Model has been extremely effective at
gleaning information about which voxels are "active" and which
are not, its purpose is only to say which regions are "statistically
significant". However, considering that voxels consist of millions
of neurons, the reality is that a
region could be active, but at some other level or time constant
than expected; and be completely missed by the linear model. 
Studies showing activation
in Cadavers of Phantoms highlight the inherant problem with the 
General Linear Model: a problem that mandates multiple
comparison testing with very high thresholds to ensure any 
sort of face validity \citep{drift}.
By moving to models that actually "learn" the 
parameters of the model, model error will be drastically
reduced allowing for much more reasonable statistical thresholds.

>>>>

\begin{document}
\section*{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
neural activity. Despite its rather limited temporal resolution, FMRI still
is the best way of measuring neural activity deep in the brain. 
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis is straight forward.
Because of these benefits, FMRI continues to have great potential for
studying human cognition. Despite the fact that FMRI is so widely used,
the choice of analysis tools is still relatively limited. Every widely
available analysis tool is based on parametric modeling, which \emph{requires}
prior knowledge of the noise, and depends on that knowledge being accurate.
This limitation casts a long shadow over any experiment that claims to be
able to reject the null hypothesis, no matter the P-value. While obviously
there is a correlation between the BOLD signal and actual activity, its
possible the current parametric tests are over-estimating the actual amount
of activity. Its worht noting that the opposite could also be true, parametric
based t-tests may in fact underestimate activity.

fact, typical activation detection does little to account for variations
in the shape of the signal, but instead is a direct T-test between the measured signal
and the expected shape of the signal. Because of these limitations,
multiple comparison tests are imperative (ref fish), and active regions
must be very large to be detected with any confidence. This paper
will introduce a more robust method of analyzing FMRI signals for 
activation, by means of estimating the BOLD parameters.

\subsection*{FMRI?}

\subsection*{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains, for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>, generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events

inherently alters the local ratio of Oxygenated Hemoglobin
to De-oxygenated hemoglobin, which of course changes the atomic makeup of the
region. Of course MRI's contrast is based on the nuclear spin excitability of
atoms, and so change in the amount of oxygen changes the local oxygen concentration of 
The changes in the FMRI signal result from the changes in the ratio
of Oxygenated Hemaglobin to Deoxygenated Hemaglobin.

\section*{Current Techniques}
\subsection*{Basic T-Test}
\subsection*{General Linear Model}

\section*{Proposed Approach}
\subsection{Modeling}
The short term goal of current neurological models is to
learn something about the system being modeled.
While control (for instance in prosthetics), or predication
(such as BlueBrain) may some day be feasible, in the
short run we still have a great deal to learn about how our
brain works. By learning about specific systems (the dynamics
of a single person's brain), we may be able to diagnose diseases,
find their causes and even someday rectify the cause. Of course
by learning about various specific systems we can also learn 
about the general way our brains work. The reason then, for 
studying the BOLD model then is
to be able to derive the various system parameters and understand
why one brain region (say a voxel) reacts in one way, when another
acts completely differently.

While the General Linear Model has been extremely effective at
gleaning information about which voxels are "active" and which
are not, its purpose is only to say which regions are "statistically
significant". However, considering that voxels consist of millions
of neurons, the reality is that a
region could be active, but at some other level or time constant
than expected; and be completely missed by the linear model. 
Studies showing activation
in Cadavers of Phantoms highlight the inherant problem with the 
General Linear Model: a problem that mandates multiple
comparison testing with very high thresholds to ensure any 
sort of face validity \citep{drift}.
By moving to models that actually "learn" the 
parameters of the model, model error will be drastically
reduced allowing for much more reasonable statistical thresholds.
\subsection{Particle Filters}

\section*{Methods}
\subsection*{Preprocessing}
After FMRI data has been acquired it is always necessary to modify the
data in some way to make different runs comparable. Because FMRI signal
levels are essentially unit-less, at the very least it is necessary to convert
the data into \% difference from the baseline. Finding the baseline is the
first hurdle in dealing with FMRI. Various methods exist for doing this,
but they are all ad-hoc. It is common to treat each stimulus pulse independently,
ignoring the long decay of the BOLD signal. By treating each input separately,
the "base" for calculating the percent difference is merely the signal level 
before the stimulus is applied. The problem with this is that the BOLD signal
can be heavily delayed, and fall times can be as long as 10 or 15 seconds,
meaning that the assumption is clearly a false one. Another popular method 
of dealing with this problem, is to put all the time series through a high
pass filter. This will of course remove the DC component of the signal, and 
some amount of so called "drift". The problem with this method is that it is
not adaptive to the input. Huge variations in drift frequencies can exist 
in a single time-series. Thus, a single cutoff frequency could miss a significant
drift component, or it could remove \emph{actual} signal, if the cutoff frequency is
set too high.

\section*{Introduction to Modeling}
